---
title: "beta_diversity_and_stats"
output: html_document
date: "2025-02-27"
---

Load in required libraries and phyloseq object.
```{r}
library(vegan)
library(phyloseq)
library(ggplot2)

load("phyloseq_file.RData")
```
Subset phyloseq object to select for Pilosa order.
```{r}
physeq_pilosa <- subset_samples(phyloseq_file, Taxonomy_Order == "Pilosa")
```
Beta diversity analysis using Bray-Curtis dissimilarity, and visualization using a PCoA plot.
```{r}
bc_dm <- distance(physeq_pilosa, method = "bray")

pcoa_bc <- ordinate(physeq_pilosa, method="PCoA", distance=bc_dm)

gg_pcoa <- plot_ordination(physeq_pilosa, pcoa_bc, color = "Climate..basic.", shape="captive_wild") +
  labs(pch="Captivity Status", col = "Köppen-Geiger Climate Class")
gg_pcoa
```
PERMANOVA to test if there are differences between samples in the Bray-Curtis distance matrix.
```{r}
# extract metadata file from physeq_pilosa
metadata_pilosa <- as(sample_data(physeq_pilosa), "data.frame")

# PERMANOVA 
adonis2(bc_dm ~ `Climate..basic.`*captive_wild, data=metadata_pilosa)
```
Re-plot beta diversity analysis with ellipses to show significant difference.
```{r}
gg_pcoa_sigdif <- plot_ordination(physeq_pilosa, pcoa_bc, color = "Climate..basic.", shape="captive_wild") +
  stat_ellipse(type = "norm")
  labs(pch="Captivity Status", col = "Köppen-Geiger Climate Class")
gg_pcoa_sigdif

ggsave("plot_pcoa_beta_div.png"
       , gg_pcoa_sigdif
       , height=4, width=6.5)
```